The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countris

### 2025-04-08 code/processing-code/processingfile.qmd
Edit manuscript

This file contains processing data for the Final Project titlled " The effects of multi environmental exposures on Tuberculosis Incidence in Asian and African countris".

@@

Independent Variables - Basic water - Safety Water - Greenhouse - PM2.5 - Basic Sanitation - Safe Sanitation - Slum - Fule_cooking

Control variables: - Human Development Index - Health Expenditure - GDP

@@

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(purrr)
install.packages("ggplot2")
library(ggplot2)

@@

```{r}
# note the use of the here() package and not absolute paths
tbi_location <- here::here("data","raw-data","tbi.csv")
tbi <- read.csv(tbi_location)

colnames(tbi)
```

@@

***Fuel for Cooking***

```{r}
cookingfuel_location <- here::here("data","raw-data","fule_cooking.xlsx")
cookingfuel <- read_excel (cookingfuel_location )

summary(cookingfuel)
skimr::skim(cookingfuel)
colnames(cookingfuel)
```

```{r}
cookingfuel <- cookingfuel %>% 
  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns

colnames(cookingfuel) <- gsub("\\[.*\\]", "", colnames(cookingfuel))  # to delete [YR.... ] in the columns

cookingfuel <- cookingfuel %>%
  rename(country = `Country Name`)

# Reshape from wide to long format
 cookingfuel <- cookingfuel %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "pro_cleanfuel") %>%
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

 saveRDS(cookingfuel, here("data", "processed-data", "processed-cookingfuel.rds"))
```

@@

***Slum Area***

```{r}
slum_location <- here::here("data","raw-data","slum.xlsx")
slum <- read_excel (slum_location)

summary(slum)
skimr::skim(slum)
```

```{r}
slum <- slum %>% 
  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns

colnames(slum) <- gsub("\\[.*\\]", "", colnames(slum))  # to delete [YR.... ] in the columns

slum <- slum %>%
  rename(country = `Country Name`)

# Reshape from wide to long format
 slum <- slum %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "pro_slum") %>%
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

saveRDS(slum, here("data", "processed-data", "processed-slum.rds"))
```

***Greenhouses***

```{r}
greenhouses_location <- here::here("data","raw-data","greenhouse.xlsx")
greenhouses <- read_excel (greenhouses_location)

summary(greenhouses)
skimr::skim(greenhouses)
```

```{r}
greenhouses <- greenhouses %>% 
  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns

colnames(greenhouses) <- gsub("\\[.*\\]", "", colnames(greenhouses))  # to delete [YR.... ] in the columns

greenhouses <- greenhouses %>%
  rename(country = `Country Name`)

# Reshape from wide to long format
 greenhouses <- greenhouses %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "total_greenhouses") %>% # Total greenhouse gas in million metric cubic ton including emissions and removals from Land Use, Land-Use Change, and Forestry (LULUCF)
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

saveRDS(greenhouses, here("data", "processed-data", "processed-greenhouses.rds"))
```

***PM2.5***

```{r}
pm25_location <- here::here("data","raw-data","pm25.xlsx")
pm25 <- read_excel (pm25_location)

summary(pm25)
skimr::skim(pm25)
```

```{r}
pm25 <- pm25 %>% 
  select(-`Series Code`, - `1990 [YR1990]`, -`Series Name`, -`Country Code`, `2023 [YR2023]`)# delete potention confusing columns

colnames(pm25) <- gsub("\\[.*\\]", "", colnames(pm25))  # to delete [YR.... ] in the columns

pm25 <- pm25 %>%
  rename(country = `Country Name`)

# Reshape from wide to long format
 pm25 <- pm25 %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "mean_exposure_pm2.5") %>% #PM2.5 air pollution, mean annual exposure (micrograms per cubic meter)
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

 saveRDS(pm25, here("data", "processed-data", "processed-pm25.rds"))
```

***GDP***

```{r}
gdp_location <- here::here("data","raw-data","gdp.csv")
gdp <- read.csv (gdp_location)

summary(gdp)
skimr::skim(gdp)
```

```{r}
gdp <- gdp %>% 
  select(-`Series.Code`, , -`Series.Name`, -`Country.Code`)# delete potention confusing columns

colnames(gdp) <-gsub("X(\\d+)..YR\\d+.", "\\1", colnames(gdp))  # to delete [YR.... ] in the columns 
gdp <- gdp %>%
  rename(country = Country.Name)

# Reshape from wide to long format
 gdp <- gdp %>%
  pivot_longer(cols = -`country`,  # All columns except 'Country' will be pivoted
               names_to = "year", 
               values_to = "gdp") %>% #US Dollar 
  mutate(year = as.integer(year)) %>%  # Convert Year to numeric
  filter(year != 2023)  # Remove rows where year is 2023

saveRDS(gdp, here("data", "processed-data", "processed-gdp.rds"))
```

@@

```{r}
#create list of dataset
datasets <- list(tbi, cookingfuel, waterbasic, safewater, slum, pm25, greenhouses, basicsanitation, safesanitation, gdp, hdi)

# Merge all datasets based on "country" and "year"
combined_df <- reduce(datasets, full_join, by = c("country", "year"))
```

@@

### Exluding slum dataset

Slum dataset is different from other dataset, the observation is every 2 years.
I want to create subset without slum data

```{r}
no_slum <- combined_df %>% 
  select(-pro_slum)

@@

no_slum <- no_slum %>%
  filter(if_all(everything(), ~ !is.na(.) & . != "" & . != ".."))
```

@@

```{r}
# Count the number of rows for each country
complete_countries <- no_slum %>%
  group_by(country) %>%
  summarise(row_count = n()) %>%
  filter(row_count == 21) %>%
  pull(country)  # Extract only countries with exactly 21 rows

# Keep only those countries in the dataset
clean_noslum <- no_slum %>%
  filter(country %in% complete_countries)

# Print first few rows to check
head(clean_noslum)

table(clean_noslum$year, clean_noslum$country) # to make sure that all countries have complete year 
dim(clean_noslum)

unique(clean_noslum$country)

@@

```{r}
clean_noslum <- clean_noslum %>%
  mutate(across(c(pro_cleanfuel, pro_basicwater, pro_safewater, pro_basicsanitation, pro_safesanitation, mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, gdp), as.numeric))

summary(clean_noslum)

saveRDS(clean_noslum, here("data", "processed-data", "clean_noslum.rds")) # save the file

@@

### Creating clean dataset inlcuding slum

Note: Slum dataset only contain observation every two years. In this dataset, I will follow the year where slum has observation. 
```{r}
# Define the required years
required_years_alldata <- c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020)

# Filter the dataset for the required years
all_data <- combined_df %>%
  filter(year %in% required_years_alldata) %>%
  group_by(country) %>%
  filter(all(required_years_alldata %in% year)) %>%
  ungroup()

all_data  <- all_data  %>%
  filter(year %in% required_years_alldata) %>%  # Keep only required years
  group_by(country) %>%
  filter(n_distinct(year) == length(required_years_alldata)) %>%  # Ensure all years exist for each country
  ungroup()

all_data <- all_data %>%
  mutate(across(where(is.character), ~ na_if(.x, ".."))) %>%  # Convert ".." to NA only for character columns
  drop_na()  # Remove rows with any NA values

# View the filtered data
print(all_data )

unique(all_data$country)
# Optionally, save the filtered data to a new CSV file
#write.csv(filtered_hdi, "filtered_hdi_selected_years.csv", row.names = FALSE)

@@

```{r}
ggplot(all_data, aes(x = year, y = tbi, group = country, color = country)) +
  geom_line(alpha = 0.7) +  # Transparency to see overlapping lines
  labs(title = "Spaghetti Plot of TBC Incidence",
       x = "Year",
       y = "TBC Incidence") +
  theme_minimal() +
  theme(legend.position = "none") 
```

```{r}
# convert variables into numeric

all_data <- all_data %>%
  mutate(across(c(pro_cleanfuel, pro_basicwater, pro_safewater, pro_slum, pro_basicsanitation, pro_safesanitation, mean_exposure_pm2.5, total_greenhouses, pro_basicsanitation, gdp), as.numeric))
summary(all_data)
str(all_data)

saveRDS(all_data, here("data", "processed-data", "clean_data_withslum.rds")) # save the file 
write.csv(all_data, here("data", "processed-data", "clean_data_withslum.csv"), row.names = FALSE)
unique(all_data$country)
```

```{r}
all_asiaafrica <- all_data %>% 
  filter(region %in% c("South-East Asia", "Africa"))

dim(all_asiaafrica)

unique(all_asiaafrica$country)

m2 <- lm(log(tbi) ~ pro_slum, data = all_asiaafrica)
summary(m2)

plot(m2)
```

```{r}
library(reshape2)  # To reshape data for ggplot

# Compute Spearman correlation matrix
cor_matrix_spearman_slum <- cor(all_asiaafrica[, c("tbi","gdp", "hdi", "pro_cleanfuel", "pro_safewater", 
                                           "pro_basicsanitation", "pro_safesanitation", "pro_slum")], 
                           method = "spearman", use = "complete.obs")

# Melt the matrix into long format for ggplot
cor_melted_slum <- melt(cor_matrix_spearman_slum)

# Plot heatmap
ggplot(cor_melted_slum, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  geom_text(aes(label = round(value, 2)), size = 4) +  # Show correlation values
  theme_minimal() +
  labs(title = "Spearman Correlation Heatmap", x = "", y = "")
```

```{r}
ggplot(all_asiaafrica, aes(x = year, y = pro_slum, group = country)) +
  geom_line() +
  facet_wrap(~ country) +  # Creates separate plots for each country
  theme_minimal()
```

>>>>

This file contains processing data for the Final Project titlled " The effects of water and sanitation on Tuberculosis Incidence in South East Asian and African countris".

@@

- Basic water, Safety Water, Basic Sanitation, Safe Sanitation, Human Development Index (Independent Variables )

@@

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(purrr)
install.packages("ggplot2")
library(ggplot2)
install.packages("viridis")
library(viridisLite)

install.packages("tidyverse")
library(tidyverse)
install.packages("rsample") 
library(rsample)

@@

```{r}
# note the use of the here() package and not absolute paths
tbi_location <- here::here("data","raw-data","tbi.csv")
tbi <- read.csv(tbi_location)
```
```{r}
colnames(tbi)
```

@@

@@

@@

```{r}
#create list of dataset
datasets <- list(tbi,  waterbasic, safewater,  basicsanitation, safesanitation, hdi)

# Merge all datasets based on "country" and "year"
combined_df <- reduce(datasets, full_join, by = c("country", "year"))

saveRDS(hdi, here("data", "processed-data", "combined-data.rds"))
```

@@

@@

combined_df <- combined_df %>%
  filter(if_all(everything(), ~ !is.na(.) & . != "" & . != ".."))
```

@@

```{r}
# Count the number of rows for each country
complete_countries <- combined_df %>%
  group_by(country) %>%
  summarise(row_count = n()) %>%
  filter(row_count == 21) %>%
  pull(country)  # Extract only countries with exactly 21 rows

# Keep only those countries in the dataset
combined_df <- combined_df %>%
  filter(country %in% complete_countries)

# Print first few rows to check
head(combined_df)

table(combined_df$year, combined_df$country) # to make sure that all countries have complete year 
dim(combined_df)

unique(combined_df$country)

@@

```{r}
write.csv(combined_df, here("data", "processed-data", "combined-df.csv"), row.names = FALSE)

@@

***Only including countries from Couth East Asia and African COuntries***

```{r}
asia_africa <- combined_df %>% 
  filter(region %in% c("South-East Asia", "Africa"))

dim(asia_africa)

unique(asia_africa$region)

# save the asia africa dataset

write.csv(combined_df, here("data", "processed-data", "asia_africa.csv"), row.names = FALSE)
saveRDS(combined_df, here("data", "processed-data", "asia_africa.rds"))
```

***Visualize TBI*** 
```{r}

# Line plot of TBI over time by country
ggplot(asia_africa, aes(x = year, y = tbi, group = country, color = country)) +
  geom_line(alpha = 0.7) +
  labs(title = "TBI Trends in Asia and Africa (2000–2020)",
       x = "Year",
       y = "TBI") +
  theme_minimal() +
  theme(legend.position = "none")  # Hide legend if there are many countries

@@

It can be seen that Lesotho has very high observation and potentially become outlier. It is better to drop the observation.

```{r}
seaa <- asia_africa%>%
  filter(country != "Lesotho")

#save the final dataset 
write.csv(seaa, here("data", "processed-data", "seaa.csv"), row.names = FALSE)
saveRDS(seaa, here("data", "processed-data", "seaa.rds"))
```

```{r}
# Line plot of TBI over time by country
ggplot(seaa, aes(x = year, y = tbi, group = country, color = country)) +
  geom_line(alpha = 0.7) +
  labs(title = "TBI Trends in Asia and Africa (2000–2020)",
       x = "Year",
       y = "TBI") +
  theme_minimal() +
  theme(legend.position = "none")  # Hide legend if there are many countries
```

ggplot(asia_africa, aes(x = year, y = tbi, group = country, color = country)) +
  geom_line(alpha = 0.9) +
  facet_wrap(~ region) +
  scale_color_viridis_d(option = "turbo") +  # "plasma", "cividis", or "turbo" are good options
  labs(title = "TBI Trends (2000–2020) by Region",
       x = "Year",
       y = "TBI",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "right",
        legend.key.height = unit(0.5, "cm"),
        legend.text = element_text(size = 8))

# Machine Learning Data 
I will create machine learning data

```{r}
# create setseed for reproductibility 
ml_seed <- set.seed(2345)

# Split dataset into a 75% train set and a 25% test set

splitted_ml <- initial_split(asia_africa, prop= .75)
df_train <- training(splitted_ml)
df_test <- testing(splitted_ml)

# Save training set
saveRDS(df_train, file = here("data", "processed-data", "df_train.rds"))

# Save test set
saveRDS(df_test, file = here("data", "processed-data", "df_test.rds"))

### 2025-04-08 products/manuscript/Manuscript.qmd
Edit manuscript

The study applied Generalized Additive Mixed Model (GAMM). The model can
capture mixed effect of countries on Tuberculosis Incidence with non
linear relationship. The data analysis started by looking at the
relationship between Tuberculosis Incidence and Predictors. Non linear
relationship was discovered.Next step is to perform non-linear
relationship to looking at form of relationship. Human Development Index
(HDI), Access to basic water and well-managed water show curvature
relationship, while access to basic sanitation and well-managed
sanitation show polynomial. Data transformation was performed based.
After data transformation, we fit model1, which only includes water and
sanitation variables (as main variables in this study), model2 includes
all predictors (HDI, access to water and sanitation), model3 only
includes significant variables in the previous model. Model 1, 2 and 3
use Generalized Linear Mixed Model (GLMM). We also performed VIF test to
look at multicollinearity. Model 4 uses Generalized Additive Mixed
Models. Moreover, BIC and AIC tests were performed to look at the best
model. BIC and AIC tests found that Model 4 (GAMM) performs better with
the smallest BIC and AIC. To be able to interpret GAMM better, we
visualize the result into plots.

@@

There are some exploratory data analysis conducted and resulted table,
and charts. Table 1 shows the summary statistics including min, mean,
max, and standard deviation of all variables.

NOTE: I still have issue in saving the tables. I am still working on it.
I will show it next step.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS (here("code", "eda-code", "tables", "summary_table.rds"))
knitr::kable(resulttable)
```

The line graph ( @fig-1 ) shows the trend of all variables over the
period in each country. It can be seen that the majority of country
showed the decreases trend in tuberculosis incidence over 21 years
period. All countries showed the increase trend in percentage of
population using using basic and safe drinking water, basic and safe
sanitation, except Zimbabuwe showing the decrease in percentage of
population using basic drinking water and safe sanitation.

@@

#| label: fig-1
#| fig-cap: "Trends of all variable over the period in each country"
#| echo: FALSE
knitr::include_graphics(here("code","eda-code","figures","line-graph-tbivsfactors.png"))

( @fig-2 ) shows spaghetti plots in each variables. We can see the
trends of each country in every variables. It can be seen that generally
most of the countries show the increase trends in water and sanitation,
and human development index

@@

#| label: fig-2
#| fig-cap: "Spaghetti plots in each variables"
#| echo: FALSE
knitr::include_graphics(here("code","eda-code","figures","spageti.png"))

Now lets take a more detail tuberculosis incidence in each country over
the year using the following box plot. Lesotho leads the incidence
followed by Myanmar and Tanzania.

@@

#| label: fig-3
#| fig-cap: "Boxplot plots for each Country"
#| echo: FALSE
knitr::include_graphics(here("code","eda-code","figures","boxplot.png"))

As first step of statistical analysis. Using scatter plot to show the
linearity between tuberculosis incidence and predictors. ( @fig-4 )
shows the scatter plots. The scatter plots do not show linear
relationship between tuberculosis Incidence and predictors.

#| label: fig-4
#| fig-cap: "Scatter plots between Tuberculosis Incidence and Predictors "
#| echo: FALSE
knitr::include_graphics(here("code","eda-code","figures","scatterplot.png"))

Before going further, we need to look at correlation between
tuberculosis Incidence and all predictors.

Since they show non linearity, spearman's correlation performed and
showing in the following correlation matrix. See ( @fig-5 )

#| label: fig-5
#| fig-cap: "Spearman Correlation Matrix "
#| echo: FALSE
knitr::include_graphics(here("code","eda-code","figures","correlation_matrix.png"))

Now, we want to assess non linear relationship.

#| label: fig-6
#| fig-cap: "Nonlinear relationship"
#| echo: FALSE
knitr::include_graphics(here("code","analysis-code","figures","nonlinear.png"))

After performing nonlinear model, data transformation used based on the
shave or distribution of the plots from nonlinear relationship plots (
@fig-6 ) above. ( @fig-7 ) shows the data distribution after data
transformation.

#| label: fig-7
#| fig-cap: "transformed data plots"
#| echo: FALSE
knitr::include_graphics(here("code","analysis-code","figures","transformed.png"))

@@

There are four models build; fitting model with only water and
sanitation variables, then fitting models including co-variate (HDI).
The third model is to fit only significant variables. The first three
models use Generalized Linear Mixed Models (GLMM), while the last model
used Generalized Additive Mixed Models (GAMM). Those models were
compared using AIC and BIC criteria to select the best/ the final
models.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Model Criteria"
#| echo: FALSE
resulttable2=readRDS (here("code", "analysis-code", "tables", "criteria_output.rds"))
knitr::kable(resulttable2)
```

Based on BIC and AIC criteria, the best model fit the data is
Generalized Additive Mixed Models.

\$ \log(\text{TBI}) \sim s(\text{hdi}) + s(\text{pro\_basicwater}) +
s(\text{pro\_safewater}) + s(\text{pro\_basicsanitation}) +
s(\text{country}, \text{bs}="re") \$

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Generalized Additive Mixed Models"
#| echo: FALSE
resulttable3=readRDS (here("code", "analysis-code", "tables", "summary_Gamm.rds"))
knitr::kable(resulttable3)
```

This study employed a Generalized Additive Model (GAM) with a Gaussian
family and identity link function to examine the association between
tuberculosis incidence and key environmental and socioeconomic factors
across multiple countries. The model incorporated smooth functions for
continuous predictors and a random effect for country to account for
within-country variability. The results indicate that HDI and basic
sanitation are significant determinants of tuberculosis incidence, while
access to basic and safe water did not show strong associations.

The smooth function for Human Development Index (HDI) was highly
significant (p \< 2e-16), with an estimated effective degrees of freedom
(EDF) = 4.283, confirming a non-linear relationship between HDI and
tuberculosis incidence. The effect was strongest at lower HDI values,
suggesting that improvements in socioeconomic conditions may have the
greatest impact in reducing tuberculosis burden in low-HDI countries.

Among the water and sanitation variables, basic water access was not
statistically significant (p = 0.73379), indicating no strong evidence
of an association between water coverage and tuberculosis incidence.
Similarly, safe water access was not significant at α = 0.05 (p =
0.05605), suggesting only a weak relationship. However, basic sanitation
access was statistically significant (p = 0.00542, EDF = 3.845),
suggesting that higher access to sanitation is associated with a
reduction in tuberculosis incidence. This highlights the importance of
improving sanitation infrastructure as a potential intervention for
tuberculosis control.

The random effect of country (s(country)) was highly significant (p \<
2e-16, EDF = 16.768), indicating substantial between-country variation
in tuberculosis incidence. This justifies the inclusion of random
effects to account for country-level heterogeneity and underscores the
need for context-specific policy interventions tailored to national
conditions.

The model fit was strong, with an adjusted R² of 0.974, indicating that
97.4% of the variance in tuberculosis incidence was explained by the
model. The scale estimate was 0.014598, suggesting a low residual
variance, which reflects a well-fitting model.

```{r}
#| label: fig-7
#| fig-cap: "Generalized Additive Mixed Models"
#| echo: FALSE
knitr::include_graphics(here("code","analysis-code","figures","gam_diagnostics.jpg"))

@@

@@

>>>>

The study applied Generalized Additive Mixed Model (GAMM). The model can
capture mixed effect of countries on Tuberculosis Incidence with non
linear relationship. The data analysis started by looking at the
relationship between Tuberculosis Incidence and Predictors. Non linear
relationship was discovered. Human Development Index
(HDI), Access to basic water and well-managed water show curvature
relationship, while access to basic sanitation and well-managed
sanitation show polynomial. There are several models fitted in the study, we start with linear model with all predictors included, followed by Generalized Addictive Models (GAM), Generalized Mixed Addictive Models (GAMM) without interaction, Generalized Mixed Addictive Models (GAMM) with two factor interaction, and Generalized Mixed Addictive Models (GAMM) with two factor interaction after drop interaction which are not significant. After fitting the models, RMSE and AIC were printed to chose the best models. Based on the RMSEs and AICs, GAMM with interaction after deleting unsigniuficant interactions performed best. The RMSE and AIC were visulized using plot for better understanding.

To maximize test the performance of the model, machine learning was performed. We created splitted data with 75% train data and 25 % test data and setseed (2345). RMSEs and AICs were printed to see the performance of the models and plotted. The final model was fitted into a machine learning, XGBoost. The method is very plexible and can deal with the data conditions (non-linear relationship, some multicolinearity, and binomial distribution). Finally, the model was tunned.

@@

Firs of all, the pattern of the data was explored. The trend of Tuberculosis Incidence for each country during the period was assesed using spagethi plots. The ( @fig-1 ) shows that most countries shows a dicrease of tuberculosis incidence over the period of time.

@@

#| label: fig-1
#| fig-cap: "Trends of Tuberculosis of each country from 2000 to 2020"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "eda", "tbi_spageti.png"))
```

Figure ( @fig-2 ) shows the trend of percentage of population using safe water, basic water, safe sanitation, basic sanitation, tuberculosis incidence, and human development index over the prerion for each country. It can be seen that the percentage of population having access to water and sanitation and human development index increase over the time, while the tuberculosis incidence decrease. Figure ( @fig-3 ) show the trends of all variables based on region.

```{r}
#| label: fig-2
#| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars.png"))
```

```{r}
#| label: fig-3
#| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars_region.png"))
```

As first step of statistical analysis. linearity test was performed, the result shows that there is non-linear relationship between tuberculosis incidence and other predictors. Loess method is used to show the non-linear relationship betwen tuberculosis incidence and predictors. ( @fig-4 ) shows the relationship between tuberculosis incidence and predictors and ( @fig-5 ) show the relationhsip after data transformation. It can be seen that after data transformed, non-linear relationship stil exists. Therefore, Generalized Addittive Mixed Model is used in this study.

@@

#| label: fig-4
#| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plots_tbi_vs_allvars.png"))

```
```{r}
#| label: fig-5
#| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors after transfirmation "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plot_transformed.png"))
```

Before going further, we need to look at correlation between
tuberculosis Incidence and all predictors. Since they show non linearity, spearman's correlation performed and showing in the following correlation matrix. See ( @fig-6 ) shows  that some variables have multi collinearity. The matrix shows that water and sanitation have low correlation with tuberculosis incidence.

@@

#| label: fig-6
#| fig-cap: "Spearman Correlation Matrix "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "eda", "correlation_matrix.png"))

@@

We start with linear model with all predictors included, followed by Generalized Addictive Models (GAM), Generalized Mixed Addictive Models (GAMM) without interaction, Generalized Mixed Addictive Models (GAMM) with two factor interaction, and Generalized Mixed Addictive Models (GAMM) with two factor interaction after drop interaction which are not significant. AIC and RMSE are used to select the best model. ( @fig-7) show that  and Generalized Mixed Addictive Models (GAMM) with two factor interaction after dropin insignificant interaction performed better. GAMM with full interactions has low AIC and RMSE, but the degree of freedom shows negarive values, indicating inconsistency of the model.

#| label: fig-7
#| fig-cap: "AIC and RMSE "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plot_aic_rmse.png"))
```

***Result of Final Model***

We fitted a generalized additive mixed model (GAMM) using a negative binomial distribution with a log link to model tuberculosis incidence (TBI) across countries in South East Asia and Africa. The model incorporated smooth functions of year, access to basic and safe water and sanitation services, and the Human Development Index (HDI), as well as interaction terms between key environmental and socioeconomic indicators. A random effect was included for country to account for unobserved heterogeneity. (@tbl-1 ) shows the result of the final model. 
```{r}
#| label: tbl-1
#| tbl-cap: "Summary of GAMM Parametric and Smooth Terms"
#| echo: false
#| message: false
#| warning: false

# Load and display the saved gt table
gamm_tbl <- readRDS(here::here("results", "tables", "analysis", "gamm_summary_table.rds"))

gamm_tbl %>%
  gt(groupname_col = "type") %>%
  fmt_number(columns = where(is.numeric), decimals = 3) %>%
  tab_header(
    title = "Summary of GAMM Estimates",
    subtitle = "Parametric and Smooth Terms"
  ) %>%
  cols_label(
    term = "Term", estimate = "Estimate", std.error = "Std. Error",
    edf = "EDF", chi_sq = "Chi-square", p.value = "p-value"
  ) %>%
  fmt_missing(everything(), missing_text = "-")

@@

The model demonstrated excellent fit, explaining 99.7% of the deviance (adjusted R² = 0.997). Among the smooth terms, year (edf = 3.38, p < 0.001), access to basic water (edf = 6.52, p = 0.013), access to safe water (edf = 1.00, p < 0.001), and HDI (edf = 1.00, p < 0.001) showed significant non-linear associations with TBI. In contrast, basic and safe sanitation services were not significantly associated with TBI after accounting for other variables.

Notably, several two-dimensional interaction smooths were also statistically significant, including between basic water and HDI (p < 0.001), safe water and HDI (p < 0.001), safe sanitation and basic water (p < 0.001), and safe water and basic water (p < 0.001), suggesting complex joint effects of environmental infrastructure and development on TBI patterns. The random effect for country was highly significant, capturing substantial variation across national contexts.

These findings underscore the importance of improving access to clean water and addressing socioeconomic disparities to reduce tuberculosis burden, while also highlighting the complex interplay between infrastructure and development indicators.

### Machine Learning

To test the consistency of the model, machine learning was used.

Using Splitted / generated data shows that the model has a consistent result. (@fig-8 ) shows the observed and predicted values both original and generated data. It can be seen that both original data and generated data (both test and training data) have similiar pattern and the plots lay very closed or around the line.

```{r}
#| label: fig-8
#| fig-cap: "Predicted vs Observed TBI originalvs generated data "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plot_ob_pre_ml_vs_ori.png"))
```

Figure (@fig-9) shows the performance of machine learning (XGboost). The figure shows how well the model predicted tuberculosis incidence (TBI) by comparing the predicted values with the actual reported values for both the training and test data. The dotted line represents perfect prediction—where predicted and actual values would match exactly.

For the training data (right side), the dots are closely lined up along the diagonal line, which means the model predicted the TBI values very accurately for the data it was trained on.

For the test data (left side), most of the points are still close to the line, but some are spread out more, especially at higher TBI values. This means that while the model still performs well on new data, it slightly underestimates TBI in some cases when the actual rates are very high.

Overall, the plot shows that the model is reliable and provides accurate predictions for TBI, even on data it hasn’t seen before. This suggests the model can be useful for forecasting TB burden in similar settings.

```{r}
#| label: fig-9
#| fig-cap: "Predicted and observed value for Xgboost using train and test data"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plot_obs_pred_ml.png"))
```

***Tunning for XGBoost***

```{r}
#| label: fig-10
#| fig-cap: "Tunning Result for XGBoost"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "autoplot_xgb_tuning.png"))
```

Figure (@fig-10) shows the results of hyperparameter tuning for the XGBoost model, showing how different parameter values affected the model’s performance, measured by root mean square error (RMSE). Each dot represents one combination of hyperparameters tested during cross-validation. These results highlight that careful tuning of parameters, especially learning rate, tree depth, and sampling rate, is essential to significantly improve model accuracy. The tuning process allowed us to identify a combination of settings that minimized prediction error and led to a more reliable model.

@@

I am still working on it and read more paper to use in this discussion part. I will complete this part this week
